Male-biased sexual selection, but not sexual dichromatism, predicts speciation

### 2019-07-12 Sexual Selection and Speciation.Rmd
changes to figures

---
title: Male-biased sexual selection, but not sexual dichromatism, predicts speciation
  in birds
author: Justin G. Cally^1^, Devi Stuart-Fox^1^, Luke Holman^1^, Iliana Medina Guzmán^1^
  <br></br> <br></br> ^1^The University of Melbourne
csl: References/AmNat.csl
output:
  html_document:
    code_folding: hide
    depth: 3
    number_sections: no
    theme: cosmo
    toc: yes
    toc_float: yes
subtitle: Supplementary Material
bibliography: References/SS_Extinction.bib
---

```{r setup, message=FALSE, warning=FALSE}
library(ggplot2)
library(pander)
library(knitr)
library(tidyr)
library(mgcv)
library(diversitree)
library(repmis)
library(lme4)
library(ape)
library(geiger)
library(grid)
library(gridExtra)
library(nlme)
library(phytools)
library(brms)
library(ggridges)
library(caper)
library(purrr)
library(reshape2)
library(ggExtra)
library(grid)
library(dplyr)
library(BAMMtools)
library(gridExtra)
library(car)
library(coda)
library(MuMIn)
library(parallel)
library(HDInterval)
library(phangorn)
library(kableExtra)
library(RColorBrewer)
library(tibble)
library(stargazer)
library(ggraph)
library(stringr)
#devtools::install_github("Ax3man/phylopath")
library(phylopath)
library("EBImage") # for images
library("ggtree")
source("bamm_extinction/functions/check_and_fix_ultrametric.R")
source("functions/essim.R")
source("functions/diversification_rate_calculation.R")
source("functions/Vdodge_function.R")
source('functions/symmetrise_scale.R')
source('functions/Correlation_matrix.R')
source('functions/gheatmap2.R') #Gets rid of spaces between tiles
knitr::opts_chunk$set(cache=TRUE)
```

#Compiling the data

@@

##Speciation Rate

@@

##Sexual dichromatism and Male-biased Sexual Selection

@@

##Environmental predictors

@@

```{r, eval = FALSE}
#Increse the iterations (defaukt is 4) so we can obtain complete samples of each range
bird.points <- lapply(bird.ranges, 
                       function(x) {spsample(x, n=1000, type="random", iter = 30)})
```

**2)** Extracted the bioclim or other (*e.g.* NPP) data from each of the points:
```{r, eval = FALSE}
bird.values <- lapply(bird.points, function(x) {raster::extract(bioclim_data, x)})
bird.values <- as.list(data.frame((bird.values)))
```

@@

```{r, eval = FALSE}
#Obtain means per variable per species
bird.frame <- lapply(bird.values, function(x) {as.data.frame(x)})
bird.summary <- lapply(bird.frame, function(x) {
(as.data.frame(apply(x,2,mean, na.rm =T)))})

@@

#Now add column of species name: Same order carried through
bird.means <- cbind.data.frame(shps.jetz, bird.means)

@@

```{r, eval = FALSE}
bird.range.size <- sapply(bird.ranges,
                       function(x) {(area(x))})
bird.range.size <- sapply(bird.range.size, function(x) {sum(x)})
bird.range.size <- as.data.frame(bird.range.size)
bird.range.size <- cbind.data.frame(shps.jetz, bird.range.size)
rownames(bird.range.size) <- NULL
colnames(bird.range.size) <- c("binomial", "range.size.m2")

@@

#Generating biologically relevant predictors

@@

These predictors can be obtained from the compiled datasets seen here:

```{r, warning=FALSE}
plumage.scores <- read.csv('data/plumage_scores.csv')
#Generate sexual dichromatism score: 
plumage.scores$SDi <- abs(plumage.scores$Male_plumage_score - plumage.scores$Female_plumage_score)

#Make DF with enviro variables and plumage scores
complete.dataframe <- read.csv('data/complete.dataframe.csv')
complete.dataframe <- left_join(plumage.scores %>% dplyr::select(binomial, Male_plumage_score, Female_plumage_score, SDi), complete.dataframe, by = "binomial")
```

##Environmental Spatial Variability PCAs

@@

**Table S1:** Loadings for the first two PCs, from a PCA on the variation (standard error, where n ~ 1,000) of each bioclim variable except the standard deviation of temperature seasonality (sd.bioclim4) and the standard error of precipitation seasonality (sd.bioclim15), as these are already measures of variability and based on other bioclimatic variables.
```{r}
restricted.data <- complete.dataframe %>% drop_na(bioclim1) #Drop species without environmental data
PCA.bioclim <- prcomp(restricted.data[c('sd.bioclim1', 'sd.bioclim2', 'sd.bioclim3', 'sd.bioclim5', 'sd.bioclim6', 'sd.bioclim7', 'sd.bioclim8', 'sd.bioclim9', 'sd.bioclim10', 'sd.bioclim11', 'sd.bioclim12', 'sd.bioclim13', 'sd.bioclim14', 'sd.bioclim16', 'sd.bioclim17', 'sd.bioclim18', 'sd.bioclim19')], #Select sd.bioclim
                      scale = TRUE, center = TRUE)
PCA.predictions <- predict(PCA.bioclim)
restricted.data <- cbind(restricted.data, PCA.predictions)

@@

```{r, fig.width=10, fig.height=5}
#run GAM of association between range size and spatial climatic variation
PC1.gam <- mgcv::gam(PC1*(-1) ~ s(log(range.size.m2)), data = restricted.data, family = "gaussian") #Inverse as the PC1 loads to the negative (COUNTER-INTUITIVE)

@@

##Long term climate variability (LIG anomaly)

@@

```{r}
historical.variation.data <- as.data.frame(restricted.data[1])

@@

```{r, fig.width=8, fig.height=5}
#Now we can predict the PCA results: 
PCA.LIG.predictions <- as.data.frame(predict(PCA.LIG.bioclim)*(-1)) #So that higher numbers mean more variation
PCA.LIG.predictions <- rename(PCA.LIG.predictions, PC1.LIG = PC1,
       PC2.LIG = PC2)

@@

##Correlations between environmental predictors

@@

```{r, fig.width=10, fig.height=10}
#Draw plot
restricted.data$log.range.size <- log(restricted.data$range.size.m2)
pairs(restricted.data[,c("log.range.size", "bioclim4", "residuals.PC1", "PC1.LIG", "NPP")], lower.panel=panel.smooth, upper.panel=panel.cor)
```

@@

##Sexual Dimorphism

@@

#Analysis

##Simple speciation measures: Diversification rate (DR) and Node Density (ND)

@@

The functions that generate these tip rates estimates are: ``calculate.log.es``, ``calculate.nd`` and ``calculate.tb``. Here they are run on 100 trees plus a full MCC tree of passerine birds based on 2500 sample trees of the posterior [@Jetz_2012]. This was done using the ``phanghorn`` package [@phangorn_2010].

```{r}
#Obtain DR (log(es)) estimates, by calling the calculate.dr function
passerine.trees <- read.nexus('data/trees/Passerine_Trees_Full.nex')

@@

To gain an estimate of variation between trees we can take the values and obtain a mean, variance and coefficient of variation (CV). We obtain a CV value to account for unequal variation at higher estimates. The CV is expressed as the ratio of the standard deviation ($\sigma$) to the mean ($\mu$) (converting to a percentage not needed):

$C_V = \frac{\sigma}{\mu} \times 100$

Furthermore, we can account for over-dispersion by obtaining a transformed version of CV. In such a case we can calculate CV for a log-normal distribution:

${\widehat {c_{\rm {v}}}}_{\rm {ln}}={\sqrt {\mathrm {e} ^{s_{\rm {ln}}^{2}}-1}}$

Given that our _ES_ value is log-transformed, the calculation of ${\widehat {c_{\rm {v}}}}_{\rm {ln}}$ is perhaps more ideal. As for _ND_ we can see from the a ``qqnorm`` plot that value of ND is normally distributed, therefore the regular CV is more appropriate for this distribution.

```{r, eval = FALSE}
#For ND
#transpose the list so that the elements are the species
t.nd.list <- nd.list %>% purrr::transpose()
#not too sure why this works but turn the nested list into a dataframe
nd.values <- as.data.frame(t(sapply(t.nd.list, function(x) sapply(x, function(x) (x)))))
nd.values <- nd.values %>% tibble::rownames_to_column("binomial")
nd.summary <- melt(nd.values, id.vars = "binomial") %>% group_by(binomial) %>% summarise(
  mean.nd = mean(value), #use normal mean, not a rate
  var.nd = var(value), #Don't really need this if we use CV
  CV.nd = (sd(value)/mean(value)) #Given normal distribution of ND, standard CV formula is appropriate
)

@@

```{r}
dr.summary <- readRDS('data/dr.summary.rds')
#Plot a summary of logES and TB with the DRs with their weightings
dr.plot <- dr.summary %>% ggplot(aes(x = mean.loges, y = mean.nd, 
                          fill = (1/(CV.nd))/sum((1/(CV.nd)))*100, 
                      size = (1/(CV.loges))/sum((1/(CV.loges)))*100))+
  geom_point(shape = 21, alpha = 0.5)+
  theme_minimal()+
  geom_errorbarh(aes(xmin = mean.loges - 0.674*sqrt(var.loges), xmax = mean.loges + 0.674*sqrt(var.loges)), 
                 size = 0.015)+
  geom_errorbar(aes(ymin = mean.nd - 0.674*sqrt(var.nd), ymax = mean.nd + 0.674*sqrt(var.nd)), 
                size = 0.015)+
  theme(legend.position="bottom")+
  labs(size = 'Weight (%) [using DR]', y='Node Density [ND]', x= 'Log Diversification Rate [DR]', fill = 'Weight (%) [using ND] \n')+
  scale_fill_distiller(guide = "colorbar", palette = "RdPu", direction = 1, breaks=c(0,0.015,.03),
                           limits=c(0,0.03))
marginal.dr <-ggMarginal(
  p = dr.plot,
  type = 'density',
  margins = 'both',
  size = 5,
  colour = '#0000009C',
  fill = '#D12E6769'
)
```

```{r, fig.width=8, fig.height=6}
grid.newpage()
grid.draw(marginal.dr)
```
**Figure S4:** $\lambda_{DR}$ and $\lambda_{ND}$ estimates of speciation rate are correlated. Furthermore, we see similar variation across each species in the estimates across 100 trees as the inverse of the coefficient of variation (weight) is approximately equivalent across all values of ND and DR (colours and size). Notably, these estimates speciation rate are relatively uncertain across all species, with 50 % CIs in both axes plotted.

```{r, warning=FALSE, fig.width=8, fig.height=10}
restricted.data <- left_join(restricted.data, dr.summary, by = 'TipLabel')

@@

##BAMM measures of speciation and extinction

###Set up BAMM parameters

@@

```{r, eval = FALSE, results = 'hide'}
name.passerine.tree <- names(passerine.trees)

@@

###Run analysis

@@

```{r, eval = FALSE}
bamm.commands <- list()
for (x in name.passerine.tree) {
  bamm.commands[x] <- paste("bamm -c control_", x, ".txt", sep="")
}
```

@@

###Read in the event data and extract tip data

@@

```{r}
#Read in the tree and MCMC to check for convergence 
MCC.BAMM.tree  <- read.tree("data/BAMM_MCC/MCC.passerine.tre") #Same as other MCC tree already loaded
MCC.BAMM.mcmc <- read.csv( "data/BAMM_MCC/tree_MCC_mcmc_out.txt" , stringsAsFactors=F)

@@

**Table S4:** For the 100 trees that BAMM was run on effective sample size (ESS) for the two key BAMM parameters (number of evolutionary shifts and log-Likelihood) also indicates that BAMM converges with the minimum for each parameter across the 100 trees being over 200.
```{r}
ESS <- readRDS('data/ESS.rds')
summary(ESS) %>% pander()
```

@@

```{r}
# Read in Event Data
MCC.BAMM.ED <- getEventData(MCC.BAMM.tree,  "data/BAMM_MCC/tree_MCC_event_data.txt", burnin=0.1, nsamples=1000)
saveRDS(MCC.BAMM.ED, 'data/MCC.BAMM.ED.rds')
#From the Event Data we can extract

@@

```{r}
MCC.BAMM.df <- readRDS('data/MCC.BAMM.df.rds')
#Plot a summary of logES and TB with the DRs with their weightings
BAMM.MCC.plot <- MCC.BAMM.df %>% ggplot(aes(x = mean.lambda, y = mean.mu, 
                          fill = 1/var.lambda, 
                      size = 1/var.mu))+
  geom_point(shape = 21, alpha = 0.5)+
  theme_minimal()+
  geom_errorbarh(aes(xmin = mean.lambda - 0.674*sqrt(var.lambda), xmax = mean.lambda + 0.674*sqrt(var.lambda)),
                 size = 0.0025)+
  geom_errorbar(aes(ymin = mean.mu - 0.674*sqrt(var.mu), ymax = mean.mu + 0.674*sqrt(var.mu)),
                size = 0.0025)+
  # scale_y_continuous(trans = "log")+
  # scale_x_continuous(trans = "log")+
  # xlim(-8, 3)+
  # ylim(-10,2)+
  # scale_x_continuous(trans = "log")+
  # scale_y_continuous(trans = "log")+
  theme(legend.position="bottom")+
  labs(size = 'Inverse log(var) / Weight [using Lambda]', y='Log Extinction [Mu]', x= 'Log Speciation [Lambda]', fill = 'Inverse log(var) / Weight [using Mu] \n')+
  scale_fill_distiller(guide = "colorbar", palette = "Reds", direction = 1)

@@

```{r, fig.width=10, fig.height=6}
grid.newpage()
grid.draw(BAMM.variance)
```
**Figure S6:** The tip-rate estimates for BAMM are highly variable within each run of BAMM. Across most species there is high variability in the posterior distribution of tip-rate estimates. Here we show mean values, weights based on the variance and 50 % CIs.

###Analysis of BAMM results

@@

####Credible number of shifts

@@

```{r}
css <- credibleShiftSet(MCC.BAMM.ED, expectedNumberOfShifts=100, threshold=5, set.limit = 0.95)

@@

```{r} 
round(computeBayesFactors(MCC.BAMM.mcmc, expectedNumberOfShifts=100, burnin=0.1)[,1], digits = 2)
```

```{r, fig.width = 8, fig.height = 5}
plotPrior(MCC.BAMM.mcmc, expectedNumberOfShifts=100)
```
**Figure S7:** The apparent convergence of the number of shifts in the posterior is at odds with the variability seen in the CSS and although there is greater certainty in the number of shifts being within the range above, the position of those shifts remains variable.

@@

```{r, fig.width=10, fig.height=6}
load('data/Hackett_split_eventsample.rda')
css3 <- credibleShiftSet(ed, expectedNumberOfShifts=100, threshold=5, set.limit = 0.95)

@@

# BAMM.variance <- ggExtra::ggMarginal(
#   p = BAMM.MCC.plot,
#   type = 'density',
#   margins = 'both',
#   size = 5,
#   colour = 'black',
#   fill = '#BA3B1C91'
# )
```
**Figure S8:** Based on the event data from the BAMM run by @Harvey_2017 we find that in both our case and theirs the tip-rate estimates for many species are extremely variable across samples of the posterior probability. Here we present mean estimates with 50 % CIs.

@@

##PGLS Models

@@

###PGLS Models on DR and ND

@@

```{r, eval = FALSE}
#Prune tree
pruned.MCC.tree <- drop.tip(MCC.passerine,MCC.passerine$tip.label[-match(restricted.data$TipLabel, MCC.passerine$tip.label)])

@@

Inspect the $\lambda$ value, which can then be fixed for successive models.
```{r, warning=FALSE}
MCC.DR.corPagel <- readRDS('data/MCC.DR.corPagel.rds')
MCC.ND.corPagel <- readRDS('data/MCC.ND.corPagel.rds')
MCC.DR.corPagel[["modelStruct"]][["corStruct"]] %>% `names<-`("DR lambda") %>% pander()
MCC.ND.corPagel[["modelStruct"]][["corStruct"]] %>% `names<-`("ND lambda") %>% pander()
```

@@

```{r, eval = FALSE}
#Run model for DR
MCC.DR <- gls(MCC.DR ~ SDi
                         + log(range.size.m2)
                         + bioclim4 #Seasonal variation
                         + residuals.PC1 #Spatial variation
                         + PC1.LIG #Long-term climate variation
                         + NPP
                         + SDi*log(range.size.m2)
                         + SDi*bioclim4
                         + SDi*residuals.PC1
                         + SDi*PC1.LIG
                         + SDi*NPP,
                correlation = corPagel(0.985, phy = pruned.MCC.tree, fixed = TRUE), 
                data = restricted.data, 
                method = "REML")

@@

```{r, eval=FALSE}
#In both cases the top model is 1/2/3/4/5/6 no interaction terms. With no models within delta < 4:

@@

```{r, eval = F}
##########################################################
####### Read in DR models and extract estimates: #########
##########################################################

@@

##########################################################
####### Read in ND models and extract estimates: #########
##########################################################

@@

##PGLS Models on BAMM estimates

@@

```{r, eval = FALSE}
MCC.BAMM.df <- readRDS('data/MCC.BAMM.df.rds')

@@

###PGLS Model Summary

>>>>

---
title: Male-biased sexual selection, but not sexual dichromatism, predicts speciation
  in birds
author: "Justin G. Cally, Devi Stuart-Fox, Luke Holman, Iliana Medina Guzmán"
csl: References/evolution.csl
output:
  html_document:
    toc: true # table of content true
    toc_float: true # make 
    depth: 3  # upto three depths of headings (specified by #, ## and ###)
    number_sections: false  ## if you want number sections at each table header
    theme: yeti # lovely fonts and colours
    code_folding: hide # awesome buttons to show/hide the code
subtitle: Supplementary Material
bibliography: References/SS_Extinction.bib
---

```{r setup, message=FALSE, warning=FALSE}
library(ggplot2)
library(pander)
library(knitr)
library(tidyr)
library(mgcv)
library(diversitree)
library(repmis)
library(lme4)
library(ape)
library(geiger)
library(grid)
library(svglite)
library(gridExtra)
library(nlme)
library(phytools)
library(brms)
library(ggridges)
library(caper)
library(purrr)
library(reshape2)
library(ggExtra)
library(grid)
library(dplyr)
library(BAMMtools)
library(gridExtra)
library(car)
library(coda)
library(MuMIn)
library(parallel)
library(HDInterval)
library(phangorn)
library(kableExtra)
library(RColorBrewer)
library(tibble)
library(stargazer)
library(ggraph)
library(ggiraphExtra)
library(stringr)
library(Cairo)
library(gdtools)
library(data.table)
library(bindrcpp)
#devtools::install_github("Ax3man/phylopath")
#devtools::install_github("thomasp85/patchwork")
#devtools::install_github('VPetukhov/ggrastr')
library(ggrastr)
library(patchwork)
library(phylopath)
if(!requireNamespace("BiocManager", quietly = TRUE)){
    install.packages("BiocManager")
  }
#BiocManager::install(c("EBImage", "ggtree"))
library("EBImage") # for images
library("ggtree")
library("ggnewscale")
source("bamm_extinction/functions/check_and_fix_ultrametric.R")
source("functions/essim.R")
source("functions/diversification_rate_calculation.R")
source("functions/Vdodge_function.R")
source('functions/symmetrise_scale.R')
source('functions/Correlation_matrix.R')
source('functions/gheatmap2.R') #Gets rid of spaces between tiles
knitr::opts_chunk$set(cache=TRUE)
```

# Compiling the data

@@

## Speciation Rate

@@

## Sexual dichromatism and Male-biased Sexual Selection

@@

## Environmental predictors

@@

```{r spatial example one, eval = FALSE}
#Increse the iterations (defaukt is 4) so we can obtain complete samples of each range
bird.points <- lapply(bird.ranges, 
                       function(x) {spsample(x, n=1000, type="random", iter = 30)})
```

**2)** Extracted the bioclim or other (*e.g.* NPP) data from each of the points:
```{r spatial example two, eval = FALSE}
bird.values <- lapply(bird.points, function(x) {raster::extract(bioclim_data, x)})
bird.values <- as.list(data.frame((bird.values)))
```

@@

```{r spatial example three, eval = FALSE}
#Obtain means per variable per species
bird.frame <- lapply(bird.values, function(x) {as.data.frame(x)})
bird.summary <- lapply(bird.frame, function(x) {
(as.data.frame(apply(x,2,mean, na.rm =T)))})

@@

#Now add column of species name: Same order carried through
bird.means <- cbind.data.frame(bird.names, bird.means)

@@

```{r spatial example four, eval = FALSE}
bird.range.size <- sapply(bird.ranges,
                       function(x) {(area(x))})
bird.range.size <- sapply(bird.range.size, function(x) {sum(x)})
bird.range.size <- as.data.frame(bird.range.size)
bird.range.size <- cbind.data.frame(shps.jetz, bird.range.size)
rownames(bird.range.size) <- NULL
colnames(bird.range.size) <- c("binomial", "range.size.m2")

@@

# Generating biologically relevant predictors

@@

These predictors can be obtained from the compiled datasets read in here:

```{r read data, warning=FALSE}
plumage.scores <- read.csv('data/plumage_scores.csv')
#Generate sexual dichromatism score: 
plumage.scores$SDi <- abs(plumage.scores$Male_plumage_score - plumage.scores$Female_plumage_score)

#Make DF with enviro variables and plumage scores
complete.dataframe <- read.csv('data/complete.dataframe.csv')
complete.dataframe <- left_join(plumage.scores %>% dplyr::select(binomial, Male_plumage_score, Female_plumage_score, SDi), complete.dataframe, by = "binomial")

head(complete.dataframe) %>%
  kable("html") %>% 
  kable_styling() %>%
  scroll_box(width = "100%", height = "500px")
```

## Environmental Spatial Variability PCAs

@@

**Table S1:** Loadings for the first two PCs, from a PCA on the variation (standard error, where n ~ 1,000) of each bioclim variable except the standard deviation of temperature seasonality (sd.bioclim4) and the standard error of precipitation seasonality (sd.bioclim15), as these are already measures of variability and based on other bioclimatic variables.
```{r PCA bioclim}
restricted.data <- complete.dataframe %>% drop_na(bioclim1) #Drop species without environmental data
PCA.bioclim <- prcomp(restricted.data[c('sd.bioclim1', 'sd.bioclim2', 'sd.bioclim3', 'sd.bioclim5', 'sd.bioclim6', 'sd.bioclim7', 'sd.bioclim8', 'sd.bioclim9', 'sd.bioclim10', 'sd.bioclim11', 'sd.bioclim12', 'sd.bioclim13', 'sd.bioclim14', 'sd.bioclim16', 'sd.bioclim17', 'sd.bioclim18', 'sd.bioclim19')], #Select sd.bioclim
                      scale = TRUE, center = TRUE)
PCA.predictions <- predict(PCA.bioclim)
restricted.data <- cbind(restricted.data, PCA.predictions)

@@

```{r PCA GAM, fig.width=10, fig.height=5}
#run GAM of association between range size and spatial climatic variation
PC1.gam <- mgcv::gam(PC1*(-1) ~ s(log(range.size.m2)), data = restricted.data, family = "gaussian") #Inverse as the PC1 loads to the negative (COUNTER-INTUITIVE)

@@

## Long term climate variability (LIG anomaly)

@@

```{r historical data}
historical.variation.data <- as.data.frame(restricted.data[1])

@@

```{r PCA historical, fig.width=8, fig.height=5}
#Now we can predict the PCA results: 
PCA.LIG.predictions <- as.data.frame(predict(PCA.LIG.bioclim)*(-1)) #So that higher numbers mean more variation
PCA.LIG.predictions <- rename(PCA.LIG.predictions, PC1.LIG = PC1,
       PC2.LIG = PC2)

@@

## Correlations between environmental predictors

@@

```{r model predictor covariance, fig.width=10, fig.height=10}
#Draw plot
restricted.data$log.range.size <- log(restricted.data$range.size.m2)
pairs(restricted.data[,c("log.range.size", "bioclim4", "residuals.PC1", "PC1.LIG", "NPP")], lower.panel=panel.smooth, upper.panel=panel.cor)
```

@@

## Sexual Dimorphism

@@

# Analysis

## Simple speciation measures: Diversification rate (DR) and Node Density (ND)

@@

The functions that generate these tip rates estimates are: ``calculate.log.es``, ``calculate.nd`` and ``calculate.tb``. Here they are run on 1000 trees plus a full MCC tree of passerine birds based on 2500 sample trees of the posterior [@Jetz_2012]. This was done using the ``phanghorn`` package [@phangorn_2010].

```{r speciation statistics}
#Obtain DR (log(es)) estimates, by calling the calculate.dr function
passerine.trees <- read.nexus('data/trees/Passerine_Trees_Full.nex')

@@

<!-- To gain an estimate of variation between trees we can take the values and obtain a mean, variance and coefficient of variation (CV). We obtain a CV value to account for unequal variation at higher estimates. The CV is expressed as the ratio of the standard deviation ($\sigma$) to the mean ($\mu$) (converting to a percentage not needed): -->

<!-- $C_V = \frac{\sigma}{\mu} \times 100$ -->

<!-- Furthermore, we can account for over-dispersion by obtaining a transformed version of CV. In such a case we can calculate CV for a log-normal distribution:  -->

<!-- ${\widehat {c_{\rm {v}}}}_{\rm {ln}}={\sqrt {\mathrm {e} ^{s_{\rm {ln}}^{2}}-1}}$ -->

<!-- Given that our _ES_ value is log-transformed, the calculation of ${\widehat {c_{\rm {v}}}}_{\rm {ln}}$ is perhaps more ideal. As for _ND_ we can see from the a ``qqnorm`` plot that value of ND is normally distributed, therefore the regular CV is more appropriate for this distribution. -->

```{r speciation statistics wrangling, eval = FALSE}
#For ND
#transpose the list so that the elements are the species
t.nd.list <- nd.list %>% purrr::transpose()
#not too sure why this works but turn the nested list into a dataframe
nd.values <- as.data.frame(t(sapply(t.nd.list, function(x) sapply(x, function(x) (x)))))
nd.values <- nd.values %>% tibble::rownames_to_column("binomial")
nd.summary <- melt(nd.values, id.vars = "binomial") %>% group_by(binomial) %>% summarise(
  mean.nd = mean(value), #use normal mean, not a rate
  var.nd = var(value), #Don't really need this if we use CV
  CV.nd = (sd(value)/mean(value)) #Given normal distribution of ND, standard CV formula is appropriate
)

@@

```{r code for plot speciation rates RAW}
dr.summary <- readRDS('data/dr.summary.rds')
#Plot a summary of logES and TB with the DRs with their weightings
dr.plot <- dr.summary %>% ggplot(aes(x = MCC.DR, y = MCC.ND, 
                          fill = MCC.DR, 
                      size = MCC.ND))+
  geom_point(shape = 21, alpha = 0.5)+
  theme_minimal()+
  theme(legend.position="bottom")+
  labs(size = 'Node Density [ND]', y='Node Density [ND]', x= 'Diversification Rate [DR]\n (log-equal splits)', fill = 'Diversification Rate [DR]')+
  scale_fill_distiller(guide = "colorbar", palette = "RdPu", direction = 1)
marginal.dr <-ggMarginal(
  p = dr.plot,
  type = 'density',
  margins = 'both',
  size = 5,
  colour = '#0000009C',
  fill = '#D12E6769'
)
```

```{r plot speciation rates RAW, fig.width=8, fig.height=6}
grid.newpage()
grid.draw(marginal.dr)
```
**Figure S4:** $\lambda_{DR}$ and $\lambda_{ND}$ estimates of speciation rate are correlated for the MCC tree used in this study. Both of these response variables are normally distributed.

```{r speciation rates plot two, warning=FALSE, fig.width=8, fig.height=10}
restricted.data <- left_join(restricted.data, dr.summary, by = 'TipLabel')

@@

## BAMM measures of speciation and extinction

### Set up BAMM parameters

@@

```{r BAMM settings, eval = FALSE, results = 'hide'}
name.passerine.tree <- names(passerine.trees)

@@

### Run analysis

@@

```{r BAmm export, eval = FALSE}
bamm.commands <- list()
for (x in name.passerine.tree) {
  bamm.commands[x] <- paste("bamm -c control_", x, ".txt", sep="")
}
```

@@

### Read in the event data and extract tip data

@@

```{r BAMM inspection}
#Read in the tree and MCMC to check for convergence 
MCC.BAMM.tree  <- read.tree("data/BAMM_MCC/MCC.passerine.tre") #Same as other MCC tree already loaded
MCC.BAMM.mcmc <- read.csv( "data/BAMM_MCC/tree_MCC_mcmc_out.txt" , stringsAsFactors=F)

@@

**Table S4:** For the 100 trees that BAMM was run on effective sample size (ESS) for the two key BAMM parameters (number of evolutionary shifts and log-Likelihood) also indicates that BAMM converges with the minimum for each parameter across the 100 trees being over 200.
```{r ESS}
ESS <- readRDS('data/ESS.rds')
summary(ESS) %>% pander()
```

@@

```{r BAMM wrangling}
# Read in Event Data
MCC.BAMM.ED <- getEventData(MCC.BAMM.tree,  "data/BAMM_MCC/tree_MCC_event_data.txt", burnin=0.1, nsamples=1000)
saveRDS(MCC.BAMM.ED, 'data/MCC.BAMM.ED.rds')
#From the Event Data we can extract

@@

```{r BAMM plot}
MCC.BAMM.df <- readRDS('data/MCC.BAMM.df.rds')
MCC.BAMM.df$CV.lambda <- sqrt(exp(MCC.BAMM.df$var.lambda) - 1)*100
#Plot a summary of logES and TB with the DRs with their weightings
BAMM.MCC.plot <- MCC.BAMM.df %>% ggplot(aes(x = mean.lambda, y = mean.mu, 
                          fill = 1/var.lambda, 
                      size = 1/var.mu))+
  geom_point(shape = 21, alpha = 0.5)+
  theme_minimal()+
  geom_errorbarh(aes(xmin = mean.lambda - 0.674*sqrt(var.lambda), xmax = mean.lambda + 0.674*sqrt(var.lambda)),
                 size = 0.0025)+
  geom_errorbar(aes(ymin = mean.mu - 0.674*sqrt(var.mu), ymax = mean.mu + 0.674*sqrt(var.mu)),
                size = 0.0025)+
  # scale_y_continuous(trans = "log")+
  # scale_x_continuous(trans = "log")+
  # xlim(-8, 3)+
  # ylim(-10,2)+
  # scale_x_continuous(trans = "log")+
  # scale_y_continuous(trans = "log")+
  theme(legend.position="bottom")+
  labs(size = 'Inverse log(var) / Weight [using Lambda]', y='Log Extinction [Mu]', x= 'Log Speciation [Lambda]', fill = 'Inverse log(var) / Weight [using Mu] \n')+
  scale_fill_distiller(guide = "colorbar", palette = "Reds", direction = 1)

@@

```{r BAMM plot print, fig.width=10, fig.height=6}
grid.newpage()
grid.draw(BAMM.variance)
```
**Figure S6:** The tip-rate estimates for BAMM are highly variable within each run of BAMM. Across most species there is high variability in the posterior distribution of tip-rate estimates. Here we show mean values, weights based on the variance and 50 % CIs.

### Analysis of BAMM results

@@

**Credible number of shifts**

@@

```{r CSS}
css <- credibleShiftSet(MCC.BAMM.ED, expectedNumberOfShifts=100, threshold=5, set.limit = 0.95)

@@

```{r CSS two} 
round(computeBayesFactors(MCC.BAMM.mcmc, expectedNumberOfShifts=100, burnin=0.1)[,1], digits = 2)
```

```{r plot prior BAMM, fig.width = 8, fig.height = 5}
plotPrior(MCC.BAMM.mcmc, expectedNumberOfShifts=100)
```
**Figure S7:** The apparent convergence of the number of shifts in the posterior is at odds with the variability seen in the CSS and although there is greater certainty in the number of shifts being within the range above, the position of those shifts remains variable.

@@

```{r Hackett Data, fig.width=10, fig.height=6}
load('data/Hackett_split_eventsample.rda')
css3 <- credibleShiftSet(ed, expectedNumberOfShifts=100, threshold=5, set.limit = 0.95)

@@

Harvey.vs.us <- full_join(MCC.BAMM.df %>% select(TipLabel, mean.lambda), Harvey.BAMM.df %>% select(TipLabel, Harvey.lambda = mean.lambda), by = "TipLabel")
saveRDS(Harvey.vs.us, "data/Harvey.vs.us.rds")
# BAMM.variance <- ggExtra::ggMarginal(
#   p = BAMM.MCC.plot,
#   type = 'density',
#   margins = 'both',
#   size = 5,
#   colour = 'black',
#   fill = '#BA3B1C91'
# )
```
**Figure S8:** Based on the event data from the BAMM run by @Harvey_2017 we find that in both our case and theirs the tip-rate estimates for many species are extremely variable across samples of the posterior probability. Here we present mean estimates with 50 % CIs.

In comparison to $\lambda_{DR}$, the value of $\lambda_{BAMM}$ was the _mean_ drawn from a posterior distribution (*n* = 1,000) of BAMM generations, thus estimates of $\lambda_{BAMM}$ (and $\mu_{BAMM}$) have an added level of variation. To account for this variation, weights (using the inverse of the variance) were used for the PGLS models. From 1,000 posterior samples of the MCC BAMM run, the coefficient of variation (*CV*) for all log-rates of $\lambda_{BAMM}$ was relatively low (*mean CV* = `r round(mean(MCC.BAMM.df$CV.lambda), digits = 2)`, *median CV* = `r round(median(MCC.BAMM.df$CV.lambda), digits = 2)` ; *see* Figures S6 and Figure S8). Despite the convergence of the BAMM model in all runs --- where effective sample sizes of the number of shifts and log-likelihood were all greater than 200 (Table S3,Table S4) --- we found that the unique combinations of rate shifts across the large phylogenetic tree (*n* = 5,966 species) were high. This means that although the number of rate shifts reached convergence ($median = 59$; Figure S7), the locations of the rate shifts (*i.e.* the credible shift set) across the tree are highly heterogeneous. The variability in the locations for the shift configurations is a likely source of uncertainty in downstream tip-rates used as the response variable in PGLS models. Notably, the 95 % HPD interval for model estimates using BAMM rates from 100 trees was about 20-fold the 95 % confidence intervals of the estimate from the MCC tree; whereas for $\lambda_{DR}$ and $\lambda_{ND}$, the HPD 95 % interval width was equal to or less than the MCC 95 % CI (Table S9).

@@

## PGLS Models

@@

### PGLS Models on DR and ND

@@

```{r MCC DR and ND, eval = FALSE}
#Prune tree
pruned.MCC.tree <- drop.tip(MCC.passerine,MCC.passerine$tip.label[-match(restricted.data$TipLabel, MCC.passerine$tip.label)])

@@

Inspect the $\lambda$ value, which can then be fixed for successive models.
```{r MCC DR and ND lambda, warning=FALSE}
MCC.DR.corPagel <- readRDS('data/MCC.DR.corPagel.rds')
MCC.ND.corPagel <- readRDS('data/MCC.ND.corPagel.rds')
MCC.DR.corPagel[["modelStruct"]][["corStruct"]] %>% `names<-`("DR lambda") %>% pander()
MCC.ND.corPagel[["modelStruct"]][["corStruct"]] %>% `names<-`("ND lambda") %>% pander()
```

@@

```{r MCC DR ND models, eval = FALSE}
#Run model for DR
MCC.DR <- gls(MCC.DR ~ SDi
                         + log(range.size.m2)
                         + bioclim4 #Seasonal variation
                         + residuals.PC1 #Spatial variation
                         + PC1.LIG #Long-term climate variation
                         + NPP
                         + SDi*log(range.size.m2)
                         + SDi*bioclim4
                         + SDi*residuals.PC1
                         + SDi*PC1.LIG
                         + SDi*NPP,
                correlation = corPagel(0.985, phy = pruned.MCC.tree, fixed = TRUE), 
                data = restricted.data, 
                method = "REML")

@@

```{r MCC ND DR top, eval=FALSE}
#In both cases the top model is 1/2/3/4/5/6 no interaction terms. With no models within delta < 4:

@@

```{r trycatch all models, eval = F}

#Read in DR models and extract estimates:

@@

# Read in ND models and extract estimates

@@

## PGLS Models on BAMM estimates

@@

```{r BAMM mcc dredge, eval = FALSE}
MCC.BAMM.df <- readRDS('data/MCC.BAMM.df.rds')

@@

### PGLS Model Summary
